Mixture Models for Single-Cell Assays with Applications to Vaccine Studies

### 2013-06-04 MIMOSA.tex
added tilde zi

\noindent\textbf{EM algorithm:} Given an estimate of the model parameter vector $\tilde{\boldsymbol{\theta}}=\left\{\tilde{\alpha}^{(u)},\tilde{\beta}^{(u)},\tilde{\alpha}^{(s)},\tilde{\beta}^{(s)},\tilde{w}\right\}$ and the data $\mathbf{y}$, the E step consists of calculating the posterior probabilities of differential expression, defined by
$\tilde z_{i} \equiv \mathrm{Pr}(z_i=1|\mathbf{y},\tilde{\boldsymbol{\theta}})$; see Appendix A in Supplementary Material for details.
%=\frac{\tilde{w} \cdot \mathrm{L}_1(\tilde{\alpha}^{(u)},\tilde{\beta}^{(u)},\tilde{\alpha}^{(s)},\tilde{\beta}^{(s)},|\mathbf{y}_i)}{(1-\tilde{w})\cdot\mathrm{L}_0(\tilde{\alpha}^{(u)},\tilde{\beta}^{(u)}|\mathbf{y}_i)+\tilde{w}\cdot\mathrm{L}_1(\tilde{\alpha}^{(u)},\tilde{\beta}^{(u)},\tilde{\alpha}^{(s)},\tilde{\beta}^{(s)}|\mathbf{y}_i)}.\]
The M-step then consist of optimizing the complete-data log-likelihood over $\boldsymbol{\theta}$ after replacing $z_i$ by $\tilde{z}_{i}$ in \eqref{eq:cll}. Straightforward calculations lead to
$\tilde w = \sum_i{\tilde{z_i}}/I$, but unfortunately no closed form solutions exist for the remaining parameters. We use numerical optimization as implemented in R's \textit{optim} function to estimate the remaining parameters~\citep{Ihaka:1996ud}.  We initialize the $z_{i}$'s using Fisher's exact test to assign each observation to either the null or alternative components and then use the estimated $z_i$'s to estimate the $p^{(u)}_{i}$'s and $p^{(s)}_{i}$'s and use these to set the hyper-parameters to their method-of-moments estimates.

\noindent\textbf{MCMC algorithm:} We generated realizations from the posterior distribution via MCMC algorithms \citep{Gelfand1990}. All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were performed via Gibbs sampling.
Details about the algorithms and implementation are given in Appendix B of the Supplementary Material. 
%We used the
%method of \cite{Raftery:1992vp} and Raftery \citep{Walter1996} to determine
%the number of iterations, based on a short pilot run of the
%sampler. For each dataset presented here, we calculated that no more
%than about 100,000 iterations with 50,000 burn-in iterations was sufficient to estimate standard posterior quantities. To leave some margin, we used 200,000 iterations after 50,000 burn-in iterations for each dataset explored here.

>>>>

\noindent\textbf{EM algorithm:} Given an estimate of the model parameter vector $\tilde{\boldsymbol{\theta}}=\left\{\tilde{\alpha}^{(u)},\tilde{\beta}^{(u)},\tilde{\alpha}^{(s)},\tilde{\beta}^{(s)},\tilde{w}\right\}$ and the data $\mathbf{y}$, the E step consists of calculating the posterior probabilities of differential expression, defined by
$\tilde z_{i} \equiv \mathrm{Pr}(z_i=1|\mathbf{y},\tilde{\boldsymbol{\theta}})$; see Appendix B in Supplementary Material for details.
%=\frac{\tilde{w} \cdot \mathrm{L}_1(\tilde{\alpha}^{(u)},\tilde{\beta}^{(u)},\tilde{\alpha}^{(s)},\tilde{\beta}^{(s)},|\mathbf{y}_i)}{(1-\tilde{w})\cdot\mathrm{L}_0(\tilde{\alpha}^{(u)},\tilde{\beta}^{(u)}|\mathbf{y}_i)+\tilde{w}\cdot\mathrm{L}_1(\tilde{\alpha}^{(u)},\tilde{\beta}^{(u)},\tilde{\alpha}^{(s)},\tilde{\beta}^{(s)}|\mathbf{y}_i)}.\]
The M-step then consist of optimizing the complete-data log-likelihood over $\boldsymbol{\theta}$ after replacing $z_i$ by $\tilde{z}_{i}$ in \eqref{eq:cll}. Straightforward calculations lead to
$\tilde w = \sum_i{\tilde{z_i}}/I$, but unfortunately no closed form solutions exist for the remaining parameters. We use numerical optimization as implemented in R's \textit{optim} function to estimate the remaining parameters~\citep{Ihaka:1996ud}.  We initialize the $z_{i}$'s using Fisher's exact test to assign each observation to either the null or alternative components and then use the estimated $z_i$'s to estimate the $p^{(u)}_{i}$'s and $p^{(s)}_{i}$'s and use these to set the hyper-parameters to their method-of-moments estimates.

\noindent\textbf{MCMC algorithm:} We generated realizations from the posterior distribution via MCMC algorithms. All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were performed via Gibbs sampling.
Details about the algorithms and implementation are given in Appendix B of the Supplementary Material. 
%We used the
%method of \cite{Raftery:1992vp} and Raftery \citep{Walter1996} to determine
%the number of iterations, based on a short pilot run of the
%sampler. For each dataset presented here, we calculated that no more
%than about 100,000 iterations with 50,000 burn-in iterations was sufficient to estimate standard posterior quantities. To leave some margin, we used 200,000 iterations after 50,000 burn-in iterations for each dataset explored here.
